\section{Linguística da língua de sinais}
\label{linguistica-lingua-sinais}

Segundo \citeauthor{quadros-2004-estudos-linguisticos}, a linguística é o estudo científico das línguas naturais e humanas. É uma ciência que procura desvendar os princípios independentes da lógica e da informação que determinam a linguagem humana, bem como todas as formas criativas da comunicação.
Ela busca respostas para problemas essenciais relacionados à linguagem, como: Qual a natureza da linguagem humana? Como a comunicação se constitui? Quais os princípios que determinam a habilidade dos seres humanos em produzir e compreender a linguagem?~\cite{quadros-2004-estudos-linguisticos}


% ===================================================
% \cite{quadros-2004-estudos-linguisticos}
% ===================================================
% 
% A linguística é o estudo científico das línguas naturais e humanas.
% ---
% A linguística busca desvendar os princípios independentes da lógica e da informação que determinam a linguagem humana. Tais princípios são o que há de comum nos seres  humanos que possibilitam a realização das diferentes línguas. Portanto, nesse  sentido, a teoria linguística extrapola as questões do uso.  
% ---
% A linguística é uma ciência que busca respostas para problemas essenciais relacionados à linguagem  que precisam ser explicados: Qual a natureza da linguagem humana? Como  a comunicação se constitui? Quais os princípios que determinam a habilidade dos seres humanos em produzir e compreender a linguagem? A linguística desmembra tais questões com a finalidade de explicar os problemas e  elaborar uma teoria da linguagem humana e uma teoria da comunicação. 


Os primeiros estudos linguísticos sobre as línguas de sinais foram realizadas pelo professor Dr. William C. Stokoe Jr. da Universidade de Gallaudet há cerca de sessenta anos atrás. Stokoe ficou conhecido como o pai da linguística da língua de sinais e seu trabalho teve um impacto profundo na conscientização acerca da \acrshort{asl} nos Estados Unidos e em todo o mundo, e já percorremos um longo caminho desde então~\cite{stewart-2021-barrons-asl}.

Seu primeiro artigo, publicado em 1960, é intitulado ``\textit{Sign Language Structure}`` (ou Estrutura da Língua de Sinais)~\cite{stokoe-1960-sl-structure}. Ele foi seguido pela publicação do primeiro dicionário de \acrshort{asl} em 1965, intitulado ``\textit{Dictionary of American Sign Language on Linguistic Principles}`` (ou Dicionário da Língua de Sinais Americana sobre Princípios Linguísticos)~\cite{stokoe-1965-dictionary-asl}, o qual foi compilado em parceria com dois colegas Surdos. Em 1971, Stokoe estabeleceu o Laboratório de Pesquisa em Linguística da Universidade de Gallaudet~\cite{stewart-2021-barrons-asl}.


Em seus estudos, Stokoe comprovou que a \acrshort{asl} atendia a todos os critérios linguísticos de uma língua genuína -- no léxico, na sintaxe e na capacidade de  gerar uma quantidade infinita de sentenças.
A análise de suas propriedades revelou que ela apresenta organização formal nos mesmos níveis encontrados nas línguas faladas, incluindo um nível sublexical de estruturação interna do sinal (análoga ao nível fonológico das línguas orais) e um nível gramatical (morfossintático), que especifica os modos como os sinais devem ser combinados para formarem frases e orações. Dessa forma, percebeu-se que os sinais não eram imagens, mas símbolos abstratos com uma complexa estrutura interior~\cite{stokoe-1960-sl-structure,quadros-2004-estudos-linguisticos, pereira-2011-conhecimento-alem-sinais}.


Naturalmente que o trabalho de Stokoe representou o primeiro passo em relação aos estudos das línguas de sinais. Pesquisas posteriores, feitas em grande parte com a \acrshort{asl}, mostraram, entre outras coisas, a riqueza de esquemas e combinações possíveis entre  os elementos formais que servem para ampliar consideravelmente o vocabulário básico~\cite{quadros-2004-estudos-linguisticos}.
Aos estudos de Stokoe se seguiram outros, cujo objeto eram as línguas de sinais usadas pelas comunidades de Surdos em diferentes países, como França, Itália, Uruguai,  Argentina, Suécia, Brasil e muitos outros. Essas línguas são diferentes umas das outras e independem das línguas orais utilizadas nesses países~\cite{pereira-2011-conhecimento-alem-sinais}.


\subsection{Gramática}
\label{linguistica-gramatica}

De uma forma simples, a gramática é um conjunto de regras que regem a utilização de uma língua. Ela é determinada pelo grupo de pessoas que utilizam aquela língua e compreende a sua fonologia, morfologia e sintaxe~\cite{jay-2011-dont-just-sign,quadros-2004-estudos-linguisticos}.

% Ela é determinada pelo grupo de pessoas que a utilizam e, no caso das línguas de sinais, a maneira como a comunidade Surda a utiliza é o que constitui sua gramática.
% American Sign Language tem seu próprio sistema gramatical. Isso significa que a ASL tem suas próprias regras para fonologia, morfologia e sintaxe.

Como pode-se imaginar, a língua de sinais possui um sistema gramatical próprio, que apresenta recursos visuais, espaciais e gestuais que se combinam para criar algumas estruturas sem paralelo no mundo das línguas faladas. Em particular, as qualidades espaciais de sua gramática permitem que um usuário expresse mais de um pensamento simultaneamente – algo que não pode ser duplicado nas línguas orais~\cite{jay-2011-dont-just-sign, stewart-2021-barrons-asl}.


% ASL has visual, spatial, and gestural features that combine to create some grammatical structures that are unparalleled in the world of spoken languages. 
% In particular, the spatial qualities of ASL grammar allow a signer to express more than one thought simultaneously -- a characteristic that cannot be duplicated in English. Facial grammar, or nonmanual signals, is also a significant component of ASL.
% \cite{stewart-2021-barrons-asl}


% Simplificando, a gramática é um conjunto de regras para usar uma linguagem. A gramática de uma língua inclui a fonologia, morfologia e sintaxe dessa língua.
% A gramática de uma língua é determinada pelo grupo de pessoas que a utilizam. Os principais usuários da American Sign Language são os membros da comunidade surda. A maneira como a comunidade surda usa a ASL é o que constitui a gramática da ASL.
% American Sign Language tem seu próprio sistema gramatical. Isso significa que a ASL tem suas próprias regras para fonologia, morfologia e sintaxe.
% \cite{jay-2011-dont-just-sign}

Nas seções a seguir exploraremos um pouco mais essas particularidades e discutiremos a fonologia, morfologia e sintaxe da língua de sinais.


% ======================================================================
\subsubsection{Fonologia}
\label{linguistica-gramatica-fonologia}

A fonologia é o ramo da linguística que busca identificar a estrutura e a organização dos constituintes fonológicos, propondo  modelos descritivos e explanatórios. Em outras palavras, ela estuda as maneiras pelas quais as palavras são compostas a partir de unidades menores sem significado -- os fonemas --, bem como o que são essas unidades e de que forma elas funcionam em conjunto~\cite{quadros-2004-estudos-linguisticos,hill-2019-sign-languages}.

% Phonology
% In language, phonology is the study of the smallest part of the language that conveys meaning. In spoken languages, like English, a phoneme is a unit of sound that conveys meaning. 
% In ASL, the smallest parts of the language, the phonemes, are handshape, movement, palm orientation, location, and facial expression. If you change any of these parameters of a sign, then you have changed the meaning of the sign.
% \cite{jay-2011-dont-just-sign}

%It is about the ways in which words are made up of pieces that are not meaningful. It is about what these pieces are and how they work together.
%\cite{hill-2019-sign-languages}


Em \citeyear{stokoe-1960-sl-structure}, \citeauthor{stokoe-1960-sl-structure} definiu três fonemas ou parâmetros para a língua de sinais, os quais são articulados simultaneamente para compor os sinais: configuração das mãos, localização e movimento. Mais adiante, em \citeyear{battison-1974-phono-deletion}, \citeauthor{battison-1974-phono-deletion} introduziu um quarto parâmetro, que refere-se à orientação das palmas das mãos. Estudos posteriores, como os de \citeauthor{baker-padden-1978-nonmanual-components} em \citeyear{baker-padden-1978-nonmanual-components}, incluíram traços não manuais -- como expressão facial, movimentos da boca e direção do olhar -- como distintivos na língua de sinais~\cite{stokoe-1960-sl-structure,battison-1974-phono-deletion,baker-padden-1978-nonmanual-components}.


Dessa forma, a compreensão atual da fonologia é de que os sinais são compostos a partir dos cinco parâmetros a seguir~\cite{stewart-2021-barrons-asl,jay-2011-dont-just-sign,quadros-2004-estudos-linguisticos}:

\begin{enumerate}
   \item \textbf{Configuração de mão}: é a configuração assumida pela mão ao produzir o sinal, com um ou mais dedos selecionados em uma posição específica -- estendidos, fechados, curvados, ou dobrados. A configuração da mão pode permanecer estática ou variar na articulação do sinal, e é possível que ambas mãos apresentem uma configuração igual ou não nesse processo. A \autoref{fig:config-mao-comuns-asl} ilustra algumas configurações de mãos comumente utilizadas na \acrshort{asl}.
   
    \imagem
        {fig:config-mao-comuns-asl} % Label
        {imagens/fundamentacao/configuracoes_mao_asl} % Path
        {width=0.5\linewidth} % Size
        {Configurações de mãos comumente utilizadas na \acrshort{asl}.} % Caption
        {jay-2011-dont-just-sign} % Citation

    % is the shape of the hands when the sign is formed. The handshape may remain the same throughout the sign or it can change. If two hands are use to make a sign, both hands can have the same handshape or be different.

    % Esta é a forma da sua mão que é usada para criar o sinal.
    % This is the shape of your hand that is used to create the sign. \cite{jay-2011-dont-just-sign,hill-2019-sign-languages}

    %the configuration assumed by the hand while producing the sign, with one or more selected fingers in a particular position – extended, closed, curved, or bent.

    % Todos os sinais são formados usando um formato de mão específico. Se você alterar a forma de mão de um sinal, poderá alterar o significado do sinal. Por exemplo, se você mudar o formato de mão do sinal CIÊNCIA para um formato de mão B, você estaria assinando o sinal inicializado BIOLOGIA. Portanto, é importante saber como formar com precisão a(s) forma(s) de mão de cada signo.
    % All signs are formed using a specific handshape. If you change the handshape of a sign, you can change the meaning of the sign. For example, if you change the handshape of the sign SCIENCE to a B-handshape, you would be signing the initialized sign BIOLOGY. So, it is important to know how to accurately form the handshape( s) of each sign. \cite{jay-2011-dont-just-sign}



    \item \textbf{Orientação}: é a direção para a qual a palma da mão aponta durante a articulação do sinal. Por exemplo, a palma pode estar voltada para o corpo ou para fora, para o chão ou para cima, ou ainda assumir uma orientação distinta da outra mão. A \autoref{fig:orientacoes} ilustra as seis orientações que as mãos podem assumir.
    
    \imagem
        {fig:orientacoes} % Label
        {imagens/fundamentacao/orientacoes} % Path
        {width=0.5\linewidth} % Size
        {Orientações assumidas pela palma da mão na articulação de um sinal} % Caption
        {quadros-2004-estudos-linguisticos} % Citation
    
    % is the position of the hand(s) relative to the body. For example, the palms can be facing the body or away from the body, facing the ground, or facing upward. \cite{stewart-2021-barrons-asl}

    % This is the orientation of your palm when making the sign. \cite{jay-2011-dont-just-sign} 

    % The palm orientation of a sign refers to the position of the palms of your hands and the direction they are facing. If you change the palm orientation of a sign, you can change the meaning of the sign. \cite{jay-2011-dont-just-sign} 

    % the direction the palm points while producing the sign. (cleison)


    \item \textbf{Movimento}: corresponde à direção na qual a mão se move em relação ao corpo para criar o sinal. É um parâmetro complexo porque englobar uma ampla variedade de modos e direções, desde um sutil deslizar entre as mãos, movimentos internos das mãos e punhos, até uma trajetória complexa desenhada no espaço; movimentos em círculos, para cima, para baixo, para frente ou para trás; entre outros.
    Além disso, alguns sinais podem ser realizados utilizando-se apenas uma mão, outros envolvem ambas as mãos, e outros ainda envolvem ambas mãos mas movem apenas uma delas, como ilustra a \autoref{fig:movimentos-maos}.

    \imagem
        {fig:movimentos-maos} % Label
        {imagens/fundamentacao/movimentos_maos} % Path
        {width=0.6\linewidth} % Size
        {Os sinais podem envolver uma mão (esquerda), ambas mãos (centro), ou envolver ambas mas mover apenas uma delas (direita)} % Caption
        {jay-2011-dont-just-sign} % Citation

    % it is the direction in which the hand moves relative to the body. There is a variety of movements that range from a simple sliding movement [...] to a complex movement. \cite{stewart-2021-barrons-asl}
 
    % This is the movement of the handshape that makes the sign. \cite{jay-2011-dont-just-sign}

    % The movement of a sign is the action that is used to create the sign. The movement can be in a circle, up and down, forward or backward, etc. If you change the movement of a sign, you can change the meaning of a sign. \cite{jay-2011-dont-just-sign}

    % complex parameter that may involve a wide range of modes and directions, such as the internal  movements of the hands, wrist, and directional movements of the hands in the space (cleison)

    \item \textbf{Locação}: é o local dentro do espaço de enunciação onde um sinal é articulado. O espaço de enunciação, por sua vez, é uma área que contém todos os pontos dentro do raio de alcance das mãos em que os sinais são articulados, como ilustrado na \autoref{fig:espaco-enunciacao}. Dentro desse espaço, pode-se determinar um número finito (limitado) de locações, sendo que algumas são mais exatas -- tais como a ponta do nariz --, e outras são mais abrangentes -- como a frente do tórax. Por fim, as mãos podem permanecer fixas ou serem movidas de um local para outro durante a articulação de um sinal.
    
    \imagem
        {fig:espaco-enunciacao} % Label
        {imagens/fundamentacao/espaco_enunciacao} % Path
        {width=0.35\linewidth} % Size
        {Espaço de enunciação da língua de sinais} % Caption
        {quadros-2004-estudos-linguisticos} % Citation


    % is the place in the signing space where a sign is formed. Signs can be stationary [...] or they can move from one location in the signing space to another [...]. \cite{stewart-2021-barrons-asl}

    % This is the location of the sign on or in front of your body. \cite{jay-2011-dont-just-sign}

    % The location of a sign is where you place and form the sign in your signing area. If you change the location of a sign, you can change the meaning of a sign. Some examples of locations include: • front of your body, • Your head, • Your face (forehead, eyes, temples, ear, nose, cheek, mouth, or chin), • Your neck, • Your shoulder, chest, or stomach, • Your arm or elbow, • Your waist, • Your wrist, and • Your non-dominant hand. In this case, your non-dominant hand will most likely use one of the ABCOS15 handshapes. \cite{jay-2011-dont-just-sign}

    % Signing Area (espaço de enunciação / área de articulação dos sinais)
    % Your signing area is in the shape of a pyramid starting from the top of your head, down past your shoulders, and ending in a horizontal line at your waist. Unless you are signing a formal speech for a large audience, your signs shouldn’t move outside of this area. The space in the center of the chest is called the sightline. The sightline is where you would focus your eyes on a signer. This enables you to use your peripheral vision to see the signer’s hands and facial expressions at the same time. Make sure to try to wear solid colored clothing (without designs) when signing— it is easier on the eyes.  \cite{jay-2011-dont-just-sign}

    % Na língua de sinais brasileira, assim como em outras línguas de sinais até  o momento investigadas, o espaço de enunciação é uma área que contém  todos os pontos dentro do raio de alcance das mãos em que os sinais são  articulados. 
    % Dentro desse espaço de enunciação, pode-se determinar um número finito  (limitado) de locações, sendo que algumas são mais exatas, tais como a ponta  do nariz, e outros são mais abrangentes, como a frente do tórax (Ferreira-Brito  e Langevin, 1995). \cite{quadros-2004-estudos-linguisticos}


    % the area on the body, or in the articulation space, at which or near which the sign is articulated. (cleison)

    \item \textbf{Expressões não-manuais}: consistem nas expressões faciais e movimentos corporais que são adicionados aos sinais para criar significado. 
    Expressões faciais são capazes de prover clareza ou alterar o significado de um sinal -- por exemplo, há mais de 100 movimentos da boca na \acrshort{asl} que são utilizados para transmitir um advérbio, adjetivo ou outro significado mais descritivo; o movimento das sobrancelhas, por sua vez, pode determinar o tipo de questão que deseja-se perguntar -- como na \autoref{fig:expressoes-faciais-questoes}.
    Já os movimentos corporais são essenciais quando há necessidade de, por exemplo, descrever pessoas em diferentes posições ou locais, ou narrar histórias envolvendo pessoas em diferentes papéis.

    Expressões não-manuais desempenham duas funções essenciais na gramática: marcar construções sintáticas (frases interrogativas, orações relativas, tópicos, concordância e foco) e diferenciar componentes lexicais (referências específicas, referências pronominais, partículas negativas, advérbios, grau ou aspecto). 
    
    \imagem
        {fig:expressoes-faciais-questoes} % Label
        {imagens/fundamentacao/expressoes_faciais_questoes} % Path
        {width=0.5\linewidth} % Size
        {Expressões faciais utilizadas nas perguntas do tipo ``sim/não`` (esquerda) e nos demais tipos (direita) na \acrshort{asl}} % Caption
        {jay-2011-dont-just-sign} % Citation

    
    % Body Language/ Role Shifting 
    % Body language is also one of the many non-manual markers. You would use body language for things like role shifting. Role shifting is very much fun in ASL. It’s like acting. 
    % Role shifting is when you take on the “role” of another person and show what that person said, did, or felt. Essentially, you can sign a whole story using role shifting.
    % You need to be sure to include the appropriate facial expressions and other non-manual markers when role playing each person. For example, if you are showing a conversation between a parent and a child, you would take on the characteristics of the parent and gaze downward when you shift one way, and take on the characteristics of the child and gaze upward when you shift the other way. 
    % This is also essential if you are showing people in different positions or locations. You would need to be able to show if someone is standing or sitting, laying down or kneeling, etc. For example, because you can’t turn and sign, you can show that you are talking to someone sitting behind you in a car by using eye gaze and pointing to their location before taking on their characteristics.
    % \cite{jay-2011-dont-just-sign}



    
    % Nonmanual markers add to signs to create meaning. They consist of various facial expressions, head tilting, shoulder movement, and mouth movements. With a non-manual marker, the meaning of the sign can change completely. \cite{stewart-2021-barrons-asl}

    % This is the various facial expressions or body movements that are used to create meaning. \cite{jay-2011-dont-just-sign}

    % Non-Manual Markers are very important in American Sign Language. They consist of the various facial expressions and body movements that are added to signs to create meaning. Non-manual markers can be facial expressions, head shakes, head nods, head tilts, shoulder shrugs, etc. 
    % And not only do non-manual markers have a very important role in ASL grammar, but if you do not use any non-manual markers or facial expressions, your audience may not understand what you are signing, or worse, they may get bored very quickly. In English, this would be like speaking in a monotone voice. \cite{jay-2011-dont-just-sign}

    % Facial Expressions 
    % Facial expressions are the non-manual markers that refer only to the expressions on your face. The meaning of your sign can be affected by the type of facial expression you use while signing it.
    % Facial expressions can also determine what type of question you are asking. If you raise your eyebrows while asking a question, you are asking a yes or no question. If you lower your eyebrows while asking a question, you are asking a question that requires more than a yes or no answer (generally a “wh” word question).
    % Facial expressions also add clarity to what you mean when you are signing. Some signs even require a certain facial expression in order to sign them. For example, the only difference between the signs LATE and NOT-YET is the facial expression. NOT-YET is signed with your tongue hanging out slightly over your bottom teeth. Without this facial expression, the meaning of the sign changes. And there are over 100 mouth movements in ASL that are used to convey an adverb, adjective, or another more descriptive meaning when signing certain ASL words.

    % Body Language/ Role Shifting 
    % Body language is also one of the many non-manual markers. You would use body language for things like role shifting. Role shifting is very much fun in ASL. It’s like acting. 
    % Role shifting is when you take on the “role” of another person and show what that person said, did, or felt. Essentially, you can sign a whole story using role shifting.
    % You need to be sure to include the appropriate facial expressions and other non-manual markers when role playing each person. For example, if you are showing a conversation between a parent and a child, you would take on the characteristics of the parent and gaze downward when you shift one way, and take on the characteristics of the child and gaze upward when you shift the other way. 
    % This is also essential if you are showing people in different positions or locations. You would need to be able to show if someone is standing or sitting, laying down or kneeling, etc. For example, because you can’t turn and sign, you can show that you are talking to someone sitting behind you in a car by using eye gaze and pointing to their location before taking on their characteristics.
    % \cite{jay-2011-dont-just-sign}
    

    % movements performed by the face, eyes, head, and trunk. If we consider a more granular level, this might also include mouth movements, eyebrows, cheeks, shoulders, and others. They perform two essential functions: 1) marking syntactic constructions (interrogative sentences, relative clauses, topicalizations, agreement, and focus) and 2) differentiating lexical components (specific references, pronominal references, negative particles, adverbs, degree, or aspect). (cleison)

\end{enumerate}








% \cite{stewart-2021-barrons-asl} --------------------------------------
% The Physical Dimensions of ASL: The Five Parameters
% ASL is a visual-gestural language. It is visual because we see it and gestural because the signs are formed by the hands. Signing alone, however, is not an accurate picture of ASL. How signs are formed in space is important to understanding what they mean. The critical space is called the signing space and extends from the waist to just above the head and to just beyond the sides of the body. This is also the space in which the hands can move comfortably. As you will learn in this book, the signing space has a role in ASL grammar. Two or more concepts can be simultaneously expressed in ASL. This feat cannot be accomplished in a spoken language because speech is temporal in that one word rolls off the tongue at a time. One further dimension of ASL is the movement of the head and facial expressions, which help shape the meaning of ASL sentences.
% ---
% How Are Signs Formed?
% The five parameters below come together to create a sign.
% 1) Handshape is the shape of the hands when the sign is formed. The handshape may remain the same throughout the sign or it can change. If two hands are use to make a sign, both hands can have the same handshape of be different.
% 2) Orientation is the position of the hand(s) relative to the body. For example, the palms can be facing the body or away from the body, facing the ground, or facing upward.
% 3) Location is the place in the signing space where a sign is formed. Signs can be stationary [...] or they can move from one location in the signing space to another [...].
% 4) Movement of a sign is the direction in which the hand moves relative to the body. There is a variety of movements that range from a simple sliding movement [...] to a complex movement.
% 5) Nonmanual markers add to signs to create meaning. They consist of various facial expressions, head tilting, shoulder movement, and mouth movements. With a non-manual marker, the meaning of the sign can change completely.
% ---


% \cite{jay-2011-dont-just-sign} ---------------------
% Phonology
% In language, phonology is the study of the smallest part of the language that conveys meaning. In spoken languages, like English, a phoneme is a unit of sound that conveys meaning. 
% In ASL, the smallest parts of the language, the phonemes, are handshape, movement, palm orientation, location, and facial expression. If you change any of these parameters of a sign, then you have changed the meaning of the sign.
% ---
% The Five Sign Parameters
% Just like how we see English words as the arrangement of letters, there are five basic sign language elements that make up each sign. If any of these parameters are changed when creating a sign, the meaning of the sign can change. 
% The first four elements are: 
% • Handshape – This is the shape of your hand that is used to create the sign. 
% • Movement – This is the movement of the handshape that makes the sign. 
% • Palm orientation – This is the orientation of your palm when making the sign. 
% • Location – This is the location of the sign on or in front of your body. 
% There is also a fifth element that has recently been included with this list: 
% • Non-manual Markers – This is the various facial expressions or body movements that are used to create meaning. 
% American Sign Language is a very expressive language, and understanding these elements will give you a better understanding of how signs are made and what makes them different.
% ---


% \cite{quadros-2004-estudos-linguisticos} ------------------------------
% Fonologia das línguas de sinais
% Fonologia das línguas de sinais é o ramo da lingüística que objetiva identificar a estrutura e a organização dos constituintes fonológicos, propondo  modelos descritivos e explanatórios. A primeira tarefa da fonologia para línguas de sinais é determinar quais são as unidades mínimas que formam os  sinais. A segunda tarefa é estabelecer quais são os padrões possíveis de combinação entre essas unidades e as variações possíveis no ambiente fonológico. 


% \cite{hill-2019-sign-languages} -------------------------------
% Phonology
% Phonology is often described as the study of the sounds of language and their organization. If that is the way to look at phonology, it would be appropriate to say that sign languages do not have phonology. However, phonology is actually more abstract. It is about the ways in which words are made up of pieces that are not meaningful. It is about what these pieces are and how they work together. In spoken languages, the actual sounds that make up words are part of the study of phonology; yet, phonology is more concerned with the ways we can define these component pieces, how they change in different contexts (e.g., different words), and how they are organized.
% If there are component pieces that make up individual signs, then there is sign phonology. If there are implicit rules about the ways that the pieces can and cannot combine, then there is phonology. Indeed, one of the first linguistic discoveries about American Sign Language (ASL) is that “signs have parts” – [...]

% The Five Sign Parameters
% Just like how we see English words as the arrangement of letters, there are five basic sign language elements that make up each sign. If any of these parameters are changed when creating a sign, the meaning of the sign can change. 
% The first four elements are: 
% • Handshape – This is the shape of your hand that is used to create the sign. 
% • Movement – This is the movement of the handshape that makes the sign. 
% • Palm orientation – This is the orientation of your palm when making the sign. 
% • Location – This is the location of the sign on or in front of your body. 
% There is also a fifth element that has recently been included with this list: 
% • Non-manual Markers – This is the various facial expressions or body movements that are used to create meaning. 
% American Sign Language is a very expressive language, and understanding these elements will give you a better understanding of how signs are made and what makes them different.



% ======================================================================
\subsubsection{Morfologia}
\label{linguistica-gramatica-morfologia}

A morfologia é o ramo da linguística que estuda como as palavras são formadas a partir de partes componentes. Um morfema é geralmente descrito como um par consistente de forma (por exemplo, uma sequência de sons ou uma combinação de formato de mão, localização e movimento) e significado, mas existem morfemas que mudam de forma em diferentes contextos, bem como aqueles que não mudam. t parecem ter um significado consistente. \cite{hill-2019-sign-languages}

% ---------------------------------------------------------------------
% \cite{hill-2019-sign-languages}
% Morphology
% Morphology is the branch of linguistics that studies how words are formed from component parts. A morpheme is generally described as a consistent pairing of form (e.g., a sequence of sounds or a combination of handshape, location, and movement) and meaning, but there are morphemes that change their form in different contexts as well as those that don’t seem to have a consistent meaning.


% ---------------------------------------------------------------------
% \cite{pereira-2011-conhecimento-alem-sinais}
% Aspectos morfológicos (pág 70)
% Como a língua portuguesa, a Libras conta com um léxico e  com recursos que permitem a criação de novos sinais . Contudo, diferentemente das línguas orais, em que palavras complexas  são, muitas vezes, formadas pela adição de um prefixo ou sufixo a  uma raiz, nas línguas de sinais a raiz é frequentemente enriquecida com vários movimentos e contornos no espaço de sinalização  (Klima e Bellugi, 1979).  
% Um processo bastante comum na Libras para a criação de novos sinais é o que deriva nomes de verbos, e vice-versa, por meio  da mudança no movimento . O movimento dos nomes repete e  encurta o movimento dos verbos (Quadros e Karnopp, 2004) . 
% [imagem]
% Processo semelhante é observado em ouvir e ouvinTe . O  movimento de fechar as mãos próximo ao ouvido é mais curto e  repetido em ouvinTe 
% [imagem]
% Outro processo bastante usado na Libras, na criação de novos  sinais, é a composição . Nesse processo, dois sinais se combinam,  dando origem a um novo sinal, como se pode observar em eSCoLA  e igrejA 
% [imagem]
% O sinal de eSCoLA é composto pelos sinais de CASA e eSTuDAr,  enquanto o de igrejA é composto pelo sinais de CASA e Cruz 

% A criação de novos sinais na Libras pode ser obtida, ainda, por  meio da incorporação de um argumento, de um numeral ou de  uma negação.  A incorporação de argumento é muito frequente na Libras por  causa das características visuais e espaciais da língua . O sinal de  LAvAr, por exemplo, varia de acordo com o objeto que está, foi  ou será lavado .
% [imagem]


% A incorporação de um numeral caracteriza-se pela mudança na  configuração de mão do sinal para expressar a quantidade . Assim,  por exemplo, pela mudança na configuração de mão, de 1 para 2  ou para 3, o número de meses, dias ou horas referidos muda . A  localização, a orientação e os traços não manuais permanecem os  mesmos (Quadros e Karnopp, 2004) 
% [imagem]

% A incorporação da negação é outro processo bastante produtivo na Libras e pode dar-se pela alteração do movimento do sinal, caracterizada por mudança de direção, para fora, na maioria  das vezes, com a palma da mão também para fora (Brito, 1995) .  Cabe lembrar que, nos verbos, as formas negativas são acompanhadas de meneio negativo de cabeça e expressão facial de nega-  ção, como se pode observar nos exemplos a seguir.


% Categorias gramaticais (pág 76)
% Como a língua portuguesa, a Libras organiza seus sinais em  classes, como substantivos, verbos, pronomes, advérbios, adjetivos e numerais, entre outras. Serão consideradas aqui as categorias  que apresentam especificidades na Libras decorrentes principalmente do uso do espaço .

% Verbos  
% Os verbos na Libras estão basicamente divididos em três classes (Quadros e Karnopp, 2004, pp . 116-118): simples, direcionais e espaciais .  
% • verbos simples — são verbos que não se flexionam em pessoa e número e não incorporam afixos locativos 
% • verbos direcionais (com concordância) — são verbos  que se flexionam em pessoa, número e aspecto, mas não incorporam afixos locativos 
% • verbos espaciais — são verbos que têm afixos locativos 

% Adjetivos  
% Os adjetivos são sinais que formam uma classe específica na  Libras e estão sempre na forma neutra, não recebendo marcação  para gênero (masculino e feminino) nem para número (singular e plural) . Muitos adjetivos, por serem descritivos e classificadores, expressam a qualidade do objeto, desenhando-a no ar ou  mostrando-a no objeto ou no corpo do emissor (Felipe, 2001) 

% Pronomes  
% • Pessoais - são expressos por meio dos sinais de apontar com o dedo indicador.
% No singular, o sinal para todas as pessoas é o  mesmo; o que difere é a orientação da mão . No  plural, o formato do numeral — dois, três, quatro, até nove — apontando para pessoas ou lugares a quem se faz referência é interpretado como  nós, vocês ou eles dois, três, quatro, até nove.
% • Possessivos - são expressos com a configuração de  mãos em P e seguem os mesmos princípios da expressão dos pronomes pessoais na Libras 

% Classificadores  
% Os classificadores são formas que, substituindo o nome que as  precedem, podem vir junto com o verbo para classificar o sujeito ou  o objeto que está ligado à ação do verbo (Felipe, 2001) . Para Brito  (1995), os classificadores funcionam, em uma sentença, como partes dos verbos de movimento ou de localização . O sistema de classificadores fornece um campo de representações de categoriais que  revelam o tamanho e a forma de um objeto, a animação corporal  de um personagem ou como um instrumento é manipulado (Rayman, 1999) . Morgan (2005) refere que, nas narrativas, um classificador é, muitas vezes, usado para manter a referência a objeto ou  personagem previamente mencionado por meio de um sinal .  
% Em relação às formas dos classificadores, Brito (1995) refere  que a configuração de mão em V pode ser usada para se referir a pessoas, animais ou objetos; em C, para qualquer tipo de objeto  cilíndrico, e em B, para superfícies planas, por exemplo .

% Flexão verbal  
% A flexão de número nos verbos refere-se à distinção para um,  dois, três ou mais referentes . Assim, o verbo que apresenta concordância direciona-se para um, dois ou três pontos estabelecidos  no espaço ou para uma referência generalizada incluindo todos os  referentes integrantes do discurso (Quadros e Karnopp, 2004) 
% A flexão do aspecto está relacionada com as formas e a duração dos movimentos.
% Os aspectos pontual, continuativo, durativo e iterativo são obtidos por meio de alterações do movimento e/ou da configura-  ção da mão (Brito, 1995) 
%---
%A Libras apresenta, ainda, em suas formas verbais, a marca de  tempo de forma diferente de como acontece na língua portuguesa . O tempo é marcado por meio de advérbios de tempo que  indicam se a ação está ocorrendo no presente (hoje, agora), se  ocorreu no passado (ontem, anteontem), ou se ocorrerá no futuro (amanhã, semana que vem) . Para um tempo verbal indefinido,  usam-se os sinais passado e futuro (Felipe, 2001) . Para expressar  a ideia de passado, o sinal de já, antecedendo o verbo, ou o meneio afirmativo com a cabeça, concomitante à realização do sinal,  são muito utilizados 

% Flexão nominal  
% Diferentemente da língua portuguesa na modalidade oral, que  apresenta flexão de gênero modificando os nomes, a indicação de  sexo na Libras é marcada por um sinal que indica marca de gênero feminino ou masculino, antecedendo o nome .  
% Nos substantivos, a flexão de plural é obtida, na maioria das  vezes, pela repetição do sinal, pela anteposição ou posposição de  sinais referentes aos números, ou pelo movimento semicircular, que deve abranger as pessoas ou os objetos envolvidos (Brito, 1995) 


% ---------------------------------------------------------------------
% \cite{quadros-2004-estudos-linguisticos}

% Morfologia das línguas de sinais
% Morfologia é o estudo da estrutura interna das palavras ou dos sinais,  assim como das regras que determinam a formação das palavras. A palavra  morfema deriva do grego morphé, que significa forma. Os morfemas são as  unidades mínimas de significado.  
% Alguns morfemas por si só constituem palavras, outros nunca formam  palavras, apenas constituindo partes de palavras. Desta forma, têm-se os  morfemas presos que, em geral, são os sufixos e os prefixos, uma vez que não  podem ocorrer isoladamente, e os morfemas livres que constituem palavras.  
% Mas, na medida em que se pode formar palavras a partir de outras palavras, é importante reconhecer que as palavras podem ser unidades complexas, constituídas de mais de um elemento. 
% ---
% Assim como as palavras em todas as línguas humanas, mas diferentemente dos gestos, os sinais pertencem a categorias lexicais ou a classes de palavras  tais como nome, verbo, adjetivo, advérbio, etc. As línguas de sinais têm um  léxico e um sistema de criação de novos sinais em que as unidades mínimas  com significado (morfemas) são combinadas. Entretanto, as línguas de sinais  diferem das línguas orais no tipo de processos combinatórios que freqüentemente cria palavras morfologicamente complexas. Para as línguas orais, palavras complexas são muitas vezes formadas pela adição de um prefixo ou sufixo  a uma raiz. Nas línguas de sinais, essas formas resultam freqüentemente de  processos não-concatenativos em que uma raiz é enriquecida com vários movimentos e contornos no espaço de sinalização (Klima e Bellugi, 1979). 

% [...]






% ======================================================================
\subsubsection{Sintaxe}
\label{linguistica-gramatica-sintaxe}


% \cite{pereira-2011-conhecimento-alem-sinais}
% Primeiros estudos (intro a linguística):
% As primeiras pesquisas linguísticas sobre as línguas de sinais,  mais especificamente sobre a língua de sinais americana, foram realizadas por William Stokoe, no início dos anos 1960, e tiveram como objetivo mostrar que os sinais poderiam ser vistos  como mais do que gestos holísticos aos quais faltava uma estrutura interna (Stokoe, 1960) . Ao contrário do que se poderia pensar à primeira vista, eles poderiam ser descritos em termos de um  conjunto limitado de elementos formacionais que se combinavam para formar os sinais .  
% A análise das propriedades formais da língua de sinais americana revelou que ela apresenta organização formal nos mesmos níveis encontrados nas línguas faladas, incluindo um nível sublexical de estruturação interna do sinal (análoga ao nível fonológico das línguas orais) e um nível gramatical, que especifica os modos como os sinais devem ser combinados para formarem frases e orações (Klima e Bellugi, 1979).  
% Aos estudos sobre a língua de sinais americana se seguiram outros, cujo objeto eram as línguas de sinais usadas pelas comunidades de surdos em diferentes países, como França, Itália, Uruguai,  Argentina, Suécia, Brasil e muitos outros.  
% Essas línguas são diferentes umas das outras e independem das línguas orais utilizadas nesses países 


% \cite{stewart-2021-barrons-asl}
% ASL Awareness
% Awareness of ASL has been growing since Professor William C. Stokoe, Jr., of Gallaudet University, known as the father of ASL linguistics, published his research on the linguistics of ASL about sixty years ago. His first paper, published in 1960, is titled "Sign Language Structure". This was followed by the first dictionary of ASL in 1965, Dictionary of American Sign Language on Linguistic Principles. Stokoe compiled the dictionary with two Deaf colleagues at Gallaudet, Carl Croneberg and Dorothy Casterline. In 1971, Stokoe established the Linguistic Research Laboratory at Gallaudet. Stokoe's work had a profound impact on ASL awareness in the United States and throughout the world, and we've even come a long way since then.
% ASL courses in high schools and colleges are booming. The television and movies industry has discovered the value of including Deaf actors and actresses in films. [...]

% The Physical Dimensions of ASL: The Five Parameters
% ASL is a visual-gestural language. It is visual because we see it and gestural because the signs are formed by the hands. Signing alone, however, is not an accurate picture of ASL. How signs are formed in space is important to understanding what they mean. The critical space is called the signing space and extends from the waist to just above the head and to just beyond the sides of the body. This is also the space in which the hands can move comfortably. As you will learn in this book, the signing space has a role in ASL grammar. Two or more concepts can be simultaneously expressed in ASL. This feat cannot be accomplished in a spoken language because speech is temporal in that one word rolls off the tongue at a time. One further dimension of ASL is the movement of the head and facial expressions, which help shape the meaning of ASL sentences.

% How Are Signs Formed?
% The five parameters below come together to create a sign.
% 1) Handshape is the shape of the hands when the sign is formed. The handshape may remain the same throughout the sign or it can change. If two hands are use to make a sign, both hands can have the same handshape of be different.
% 2) Orientation is the position of the hand(s) relative to the body. For example, the palms can be facing the body or away from the body, facing the ground, or facing upward.
% 3) Location is the place in the signing space where a sign is formed. Signs can be stationary [...] or they can move from one location in the signing space to another [...].
% 4) Movement of a sign is the direction in which the hand moves relative to the body. There is a variety of movements that range from a simple sliding movement [...] to a complex movement.
% 5) Nonmanual markers add to signs to create meaning. They consist of various facial expressions, head tilting, shoulder movement, and mouth movements. With a non-manual marker, the meaning of the sign can change completely.

% -----
% ASL Grammar
% ASL has visual, spatial, and gestural features that combine to create some grammatical structures that are unparalleled in the world of spoken languages. As you learn about ASL grammar, look for similarities in the world of spoken languages. [...]
% In particular, the spatial qualities of ASL grammar allow a signer to express more than one thought simultaneously -- a characteristic that cannot be duplicated in English. Facial grammar, or nonmanual signals, is also a significant component of ASL.

% Facial Grammar
% What's in a sign may not be what's in the mind. To capture the sense of what a signer is signing, you must read the signer's face and body. When you listen to someone speak, you listen not only to the words but also to how the words are spoken. The tone of the voice, the rise and fall of the pitch, the length of the pause, and the steadiness of voice are all features that you latch onto with little effort in your spoken communication.
% These traits are nonexistent in signing, but they do have parallel traits that are crucial to ASL's grammar. The raised eyebrow, the tilted head, the open mouth, hunching of the shoulders, and a sign held slightly longer than others shape the meaning of the signs that are made by the hands. We call these nonmanual signals (NMS) facial grammar, which allows you to use facial expressions, your body, and gestures to add meaning and additional information to your signing. Mastering ASL cannot occur without a mastery of facial grammar.













% ########################################################################################

% ===================================================
% \cite{pereira-2011-conhecimento-alem-sinais}
% ===================================================
%
% * Seguindo convenção proposta por James Woodward (1982), neste livro será  usado o termo “surdo” para se referir à condição audiológica de não ouvir, e o termo “Surdo” para se referir a um grupo particular de pessoas surdas que partilham  uma língua e uma cultura. 
% ---
% A língua de sinais é a língua usada pela maioria dos Surdos, na  vida diária . É a principal força que une a comunidade Surda, o  símbolo de identificação entre seus membros 
% ---
% Cada país tem sua língua de sinais, como tem sua língua na  modalidade oral. As línguas de sinais são línguas naturais, ou  seja, nasceram “naturalmente” nas comunidades Surdas. Uma vez  que não se pode falar em comunidade universal, tampouco está  correto falar em língua universal.  
% Outro aspecto a considerar é a relação estreita que existe entre  língua e cultura . As línguas de sinais refletem a cultura dos diferentes países onde são usadas, e esse é mais um argumento contra  a ideia de uma língua de sinais universal 
%---
% As línguas de sinais distinguem-se das línguas orais porque utilizam o canal visual-espacial em vez do oral-auditivo . Por esse motivo, são denominadas línguas de modalidade gestual-visual  (ou visual-espacial), uma vez que a informação linguística é recebida pelos olhos e produzida no espaço, pelas mãos, pelo movimento do corpo e pela expressão facial .  
% Apesar da diferença existente entre línguas de sinais e línguas  orais, ambas seguem os mesmos princípios com relação ao fato  de que têm um léxico, isto é, um conjunto de símbolos convencionais, e uma gramática, ou seja, um sistema de regras que rege  o uso e a combinação desses símbolos em unidades maiores .  
% ---
% Primeiros estudos (intro a linguística):
% As primeiras pesquisas linguísticas sobre as línguas de sinais,  mais especificamente sobre a língua de sinais americana, foram realizadas por William Stokoe, no início dos anos 1960, e tiveram como objetivo mostrar que os sinais poderiam ser vistos  como mais do que gestos holísticos aos quais faltava uma estrutura interna (Stokoe, 1960) . Ao contrário do que se poderia pensar à primeira vista, eles poderiam ser descritos em termos de um  conjunto limitado de elementos formacionais que se combinavam para formar os sinais .  
% A análise das propriedades formais da língua de sinais americana revelou que ela apresenta organização formal nos mesmos níveis encontrados nas línguas faladas, incluindo um nível  sublexical de estruturação interna do sinal (análoga ao nível fonológico das línguas orais) e um nível gramatical, que especifica os  modos como os sinais devem ser combinados para formarem frases e orações (Klima e Bellugi, 1979) .  
% Aos estudos sobre a língua de sinais americana se seguiram outros, cujo objeto eram as línguas de sinais usadas pelas comunidades de surdos em diferentes países, como França, Itália, Uruguai,  Argentina, Suécia, Brasil e muitos outros .  
% Essas línguas são diferentes umas das outras e independem  das línguas orais utilizadas nesses países 
% -----------------
% Concepções de surdez e de surdos 
% - Concepção clínico-patológica
% - Concepção socioantropológica (tentar adotar essa aqui)

% [Pág 30]:
% Martins (2004, p . 204-205) afirma que: “Sem língua não existem nem os surdos nem o modo de ser, cultural, surdo . Existiriam  apenas deficientes auditivos .” E segue com uma boa afirmação  em defesa da língua: “[ . . .] não é simplesmente o nível de audição  que vai definir quem é surdo ou deficiente auditivo” (op . cit .) . 

% [Pág 33]:
% Na história, constata-se que os Surdos sofreram perseguições  pelas pessoas ouvintes, que não aceitavam as diferenças e exigiam  uma cultura única por meio do modelo  ouvintista ou ouvintismo . São muitas  as lutas e histórias nas comunidades Surdas, em que o povo Surdo se une contra  as práticas dos ouvintes que não respeitam a cultura Surda (Strobel, 2008) .  
% Ainda hoje, muitos ouvintes tentam  diminuir os Surdos para que vivam isolados e tendo de assumir a  cultura ouvinte, como se esta fosse uma cultura única; ser “normal” para a sociedade significa ouvir e falar oralmente . Os ouvintes não prestam atenção aos Surdos que se comunicam por  meio da Libras . Consequentemente, não acreditam que os Surdos sejam capazes de estudar em faculdade ou realizar mestrado e  doutorado, por exemplo . “Os sujeitos ouvintes veem os sujeitos  surdos com curiosidade e, às vezes, zombam por eles serem diferentes” (Strobel, 2008, p . 22) .  
% A luta dos Surdos tem conduzido a várias vitórias, como o reconhecimento da Libras, o direito a tradutores e intérpretes da  língua brasileira de sinais–língua portuguesa e a uma educação  bilíngue para as crianças Surdas, que contemple a Libras e o português, este na modalidade escrita, entre muitas outras conquistas 

% [Pág 34] Cultura Surda
% Os Surdos constituem uma comunidade linguística minoritária, cujos elementos identificatórios são a língua de sinais e uma  cultura própria eminentemente visual . Têm um espírito gregário  muito importante que se manifesta em vários espaços . Esses espa-  ços “dos Surdos” são associações e clubes de Surdos onde desenvolvem suas próprias atividades . Constituem refúgios naturais da  língua de sinais e da identidade Surda (Strobel, 2008, p . 45) .  
% Diante da comunidade majoritariamente ouvinte, as comunidades Surdas apresentam suas próprias condutas linguísticas e seus  valores culturais . A comunidade Surda tem uma atitude diferente diante do déficit auditivo, já que não leva em conta o grau de  perda auditiva de seus membros . Pertencer à comunidade Surda pode ser definido pelo domínio da língua de sinais e pelos sentimentos de identidade grupal, fatores que consideram a surdez  como uma diferença, e não como uma deficiência .  
% Como ocorre com qualquer outra cultura, os membros das comunidades de Surdos compartilham valores, crenças, comportamentos e, o mais importante, uma língua diferente da utilizada  pelo restante da sociedade .  A língua de sinais, uma língua visual-espacial com gramática  própria, é uma das maiores produções culturais dos Surdos (Perlin, 2006) . Lane, Hoffmeister e Bahan (1996) referem que a língua de sinais tem basicamente três papéis para os Surdos: ela é símbolo da identidade social, é um meio de interação social e é  um depositário de conhecimento cultural.

% [Pág 97]
% A pessoa surda é definida como aquela que, por ter perda  auditiva, compreende o mundo e interage com ele por meio de  experiências visuais, manifestando sua cultura principalmente  pelo uso da Libras .




% Aspectos morfológicos (pág 70)
% Como a língua portuguesa, a Libras conta com um léxico e  com recursos que permitem a criação de novos sinais . Contudo, diferentemente das línguas orais, em que palavras complexas  são, muitas vezes, formadas pela adição de um prefixo ou sufixo a  uma raiz, nas línguas de sinais a raiz é frequentemente enriquecida com vários movimentos e contornos no espaço de sinalização  (Klima e Bellugi, 1979).  
% Um processo bastante comum na Libras para a criação de novos sinais é o que deriva nomes de verbos, e vice-versa, por meio  da mudança no movimento . O movimento dos nomes repete e  encurta o movimento dos verbos (Quadros e Karnopp, 2004) . 
% [imagem]
% Processo semelhante é observado em ouvir e ouvinTe . O  movimento de fechar as mãos próximo ao ouvido é mais curto e  repetido em ouvinTe 
% [imagem]
% Outro processo bastante usado na Libras, na criação de novos  sinais, é a composição . Nesse processo, dois sinais se combinam,  dando origem a um novo sinal, como se pode observar em eSCoLA  e igrejA 
% [imagem]
% O sinal de eSCoLA é composto pelos sinais de CASA e eSTuDAr,  enquanto o de igrejA é composto pelo sinais de CASA e Cruz 

% A criação de novos sinais na Libras pode ser obtida, ainda, por  meio da incorporação de um argumento, de um numeral ou de  uma negação.  A incorporação de argumento é muito frequente na Libras por  causa das características visuais e espaciais da língua . O sinal de  LAvAr, por exemplo, varia de acordo com o objeto que está, foi  ou será lavado .
% [imagem]


% A incorporação de um numeral caracteriza-se pela mudança na  configuração de mão do sinal para expressar a quantidade . Assim,  por exemplo, pela mudança na configuração de mão, de 1 para 2  ou para 3, o número de meses, dias ou horas referidos muda . A  localização, a orientação e os traços não manuais permanecem os  mesmos (Quadros e Karnopp, 2004) 
% [imagem]

% A incorporação da negação é outro processo bastante produtivo na Libras e pode dar-se pela alteração do movimento do sinal, caracterizada por mudança de direção, para fora, na maioria  das vezes, com a palma da mão também para fora (Brito, 1995) .  Cabe lembrar que, nos verbos, as formas negativas são acompanhadas de meneio negativo de cabeça e expressão facial de nega-  ção, como se pode observar nos exemplos a seguir.


% Categorias gramaticais (pág 76)
% Como a língua portuguesa, a Libras organiza seus sinais em  classes, como substantivos, verbos, pronomes, advérbios, adjetivos e numerais, entre outras. Serão consideradas aqui as categorias  que apresentam especificidades na Libras decorrentes principalmente do uso do espaço .

% Verbos  
% Os verbos na Libras estão basicamente divididos em três classes (Quadros e Karnopp, 2004, pp . 116-118): simples, direcionais e espaciais .  
% • verbos simples — são verbos que não se flexionam em pessoa e número e não incorporam afixos locativos 
% • verbos direcionais (com concordância) — são verbos  que se flexionam em pessoa, número e aspecto, mas não incorporam afixos locativos 
% • verbos espaciais — são verbos que têm afixos locativos 

% Adjetivos  
% Os adjetivos são sinais que formam uma classe específica na  Libras e estão sempre na forma neutra, não recebendo marcação  para gênero (masculino e feminino) nem para número (singular e plural) . Muitos adjetivos, por serem descritivos e classificadores, expressam a qualidade do objeto, desenhando-a no ar ou  mostrando-a no objeto ou no corpo do emissor (Felipe, 2001) 

% Pronomes  
% • Pessoais - são expressos por meio dos sinais de apontar com o dedo indicador.
% No singular, o sinal para todas as pessoas é o  mesmo; o que difere é a orientação da mão . No  plural, o formato do numeral — dois, três, quatro, até nove — apontando para pessoas ou lugares a quem se faz referência é interpretado como  nós, vocês ou eles dois, três, quatro, até nove.
% • Possessivos - são expressos com a configuração de  mãos em P e seguem os mesmos princípios da expressão dos pronomes pessoais na Libras 


% Classificadores  
% Os classificadores são formas que, substituindo o nome que as  precedem, podem vir junto com o verbo para classificar o sujeito ou  o objeto que está ligado à ação do verbo (Felipe, 2001) . Para Brito  (1995), os classificadores funcionam, em uma sentença, como partes dos verbos de movimento ou de localização . O sistema de classificadores fornece um campo de representações de categoriais que  revelam o tamanho e a forma de um objeto, a animação corporal  de um personagem ou como um instrumento é manipulado (Rayman, 1999) . Morgan (2005) refere que, nas narrativas, um classificador é, muitas vezes, usado para manter a referência a objeto ou  personagem previamente mencionado por meio de um sinal .  
% Em relação às formas dos classificadores, Brito (1995) refere  que a configuração de mão em V pode ser usada para se referir a pessoas, animais ou objetos; em C, para qualquer tipo de objeto  cilíndrico, e em B, para superfícies planas, por exemplo .

% Flexão verbal  
% A flexão de número nos verbos refere-se à distinção para um,  dois, três ou mais referentes . Assim, o verbo que apresenta concordância direciona-se para um, dois ou três pontos estabelecidos  no espaço ou para uma referência generalizada incluindo todos os  referentes integrantes do discurso (Quadros e Karnopp, 2004) 
% A flexão do aspecto está relacionada com as formas e a duração dos movimentos.
% Os aspectos pontual, continuativo, durativo e iterativo são obtidos por meio de alterações do movimento e/ou da configura-  ção da mão (Brito, 1995) 
%---
%A Libras apresenta, ainda, em suas formas verbais, a marca de  tempo de forma diferente de como acontece na língua portuguesa . O tempo é marcado por meio de advérbios de tempo que  indicam se a ação está ocorrendo no presente (hoje, agora), se  ocorreu no passado (ontem, anteontem), ou se ocorrerá no futuro (amanhã, semana que vem) . Para um tempo verbal indefinido,  usam-se os sinais passado e futuro (Felipe, 2001) . Para expressar  a ideia de passado, o sinal de já, antecedendo o verbo, ou o meneio afirmativo com a cabeça, concomitante à realização do sinal,  são muito utilizados 

% Flexão nominal  
% Diferentemente da língua portuguesa na modalidade oral, que  apresenta flexão de gênero modificando os nomes, a indicação de  sexo na Libras é marcada por um sinal que indica marca de gênero feminino ou masculino, antecedendo o nome .  
% Nos substantivos, a flexão de plural é obtida, na maioria das  vezes, pela repetição do sinal, pela anteposição ou posposição de  sinais referentes aos números, ou pelo movimento semicircular, que deve abranger as pessoas ou os objetos envolvidos (Brito, 1995) 




% ===================================================
% \cite{stewart-2021-barrons-asl}
% ===================================================
% What is ASL?
% American Sign Language, or ASL, is the language of the American Deaf Community. It is used in North America, and it is the only complete and natural sign language recognized by te Deaf community. This is a simple definition. Once it is understood that ASL maintains grammatical structure, syntax, and rules entirely separate from English grammar, we can begin to understand how to use it properly -- the way the American Deaf community intents.
% As you venture through this book, you will notice that ASL grammar is not only comprised of signed words or concepts, but it heavily involves the use of facial grammar to give information or meaning to signs. Facial grammar, including eye contact, facial expressions, eye gazing, and head movements are part of the unique grammatical structure of ASL. Because of this, and many other reasons, the visual language of ASL is fascinating for nonsigners to observe and quickly becomes a desired language to learn.

% English gloss
% ASL is an expressive and receptive language only. Because of the spatial and gestural qualities of ASL, there can be no convenient written form of ASL. What we can do is write English glosses of ASL signs. An English gloss is the best approximation of the meaning of a sign. It gives us a way of laying out ASL so that it can be studied and discussed, but is not a written form of ASL.

% Signing as a Choice of Communication
% [...] why, until recently, did so many hearing people know so little about signing? There are at least three reasons for this. First, Deaf people make up just a small fraction of the population in any area. Therefore, many hearing people never encounter a Deaf person in their through life. Second, speech is the dominant form of communication in society and gets the most attention. Third, Deaf people tend to socialize with one another and with hearing people who know how to sign.

% ASL Awareness
% Awareness of ASL has been growing since Professor William C. Stokoe, Jr., of Gallaudet University, known as the father of ASL linguistics, published his research on the linguistics of ASL about sixty years ago. His first paper, published in 1960, is titled "Sign Language Structure". This was followed by the first dictionary of ASL in 1965, Dictionary of American Sign Language on Linguistic Principles. Stokoe compiled the dictionary with two Deaf colleagues at Gallaudet, Carl Croneberg and Dorothy Casterline. In 1971, Stokoe established the Linguistic Research Laboratory at Gallaudet. Stokoe's work had a profound impact on ASL awareness in the United States and throughout the world, and we've even come a long way since then.
% ASL courses in high schools and colleges are booming. The television and movies industry has discovered the value of including Deaf actors and actresses in films. [...]

% The Physical Dimensions of ASL: The Five Parameters
% ASL is a visual-gestural language. It is visual because we see it and gestural because the signs are formed by the hands. Signing alone, however, is not an accurate picture of ASL. How signs are formed in space is important to understanding what they mean. The critical space is called the signing space and extends from the waist to just above the head and to just beyond the sides of the body. This is also the space in which the hands can move comfortably. As you will learn in this book, the signing space has a role in ASL grammar. Two or more concepts can be simultaneously expressed in ASL. This feat cannot be accomplished in a spoken language because speech is temporal in that one word rolls off the tongue at a time. One further dimension of ASL is the movement of the head and facial expressions, which help shape the meaning of ASL sentences.

% How Are Signs Formed?
% The five parameters below come together to create a sign.
% 1) Handshape is the shape of the hands when the sign is formed. The handshape may remain the same throughout the sign or it can change. If two hands are use to make a sign, both hands can have the same handshape of be different.
% 2) Orientation is the position of the hand(s) relative to the body. For example, the palms can be facing the body or away from the body, facing the ground, or facing upward.
% 3) Location is the place in the signing space where a sign is formed. Signs can be stationary [...] or they can move from one location in the signing space to another [...].
% 4) Movement of a sign is the direction in which the hand moves relative to the body. There is a variety of movements that range from a simple sliding movement [...] to a complex movement.
% 5) Nonmanual markers add to signs to create meaning. They consist of various facial expressions, head tilting, shoulder movement, and mouth movements. With a non-manual marker, the meaning of the sign can change completely.

% Cultural Importance
% ASL gives us access to Deaf culture. Learning ASL is not simply about learning another language. It is also about access. Even though we can learn something about any culture from reading about it, we acquire a deeper understanding when we can experience the culture or hear firsthand accounts from the people who are a part of the culture.
% ASL is one of the defining characteristics of the Deaf community. Although groups exist withing the community, such as the Black Deaf community and LGBTQAI+ Deaf communities. Deaf community members are bound instead by their language: ASL. To learn more about Deaf culture and tap into the resources of the Deaf community, you need a solid grasp of ASL.

% -----
% The Deaf Community
% Many hearing people view the world of the Deaf as a place where people don't hear -- where silence is a loud reminder of the difference between the two groups of people. But for Deaf people, silence is not the focus. What is important to us is that we obtain a lot of pleasure by being with other Deaf people. We relish the tales about other Deaf people's experiences in the Deaf community [...]

% Self-identification
% Deaf ("big D"):
% - identify as a culturally Deaf and part of the Deaf community
% - take pride in Deaf identity
% - may have an auditory device, such as cochlear implant, hearing aid, or FM system
% - may have a more severe hearing loss
% - use sign language as their primary source of communication
% - most likely attend a Deaf school/program
% - feel more comfortable in the Deaf world
%
% deaf ("little d"):
% - do not typically associate as members of the Deaf community
% - may have an auditory device, such as a cochlear implant, hearing aid, or FM system
% - may refer to hearing loss as a medical condition
% - may not use sign language as their primary choice of communication
% - may attend a mainstream school
% - may feel more comfortable in the hearing world
%
% Hard of hearing (HoH):
% - do not associate as members of the Deaf community
% - have hearing loss but may have residual hearing
% - possibly use an auditory device, such as a hearing or FM system, to access sounds
% - refer to hearing loss as a medical condition
% - may use sign language as their primary choice of communication
% - may attend a mainstream school
% - feel more comfortable in the hearing world

% Who Are Deaf People?
% Deaf people are a group of people who have a hearing loss, use a sign language as their primary means of communication, and have shared experiences associated with the hearing loss and the use of sign language.
% There is no way that you can point to a person sitting and reading a magazine in a lobby whom you have never met before and say, "that person is Deaf". Even if the person is wearing hearing aids, we don't know which community the person identifies with. Similarly, it's not important whether the person is European, African-American, Asian, or of some other ethnic origin. Age is not relevant, and neither is the social class or gender of the person. 
% The Deaf community is not shaped by any of these characteristics. In fact, having a hearing loss does not mean that a person is a member of the Deaf community, although it is certainly an important requirement.

% The pivotal mark of a Deaf person is how this person communicates. A Deaf person uses sign language [...]. This does not mean that the person cannot use other forms of communication, such as writing and speaking. Rather, ASL is the linguistic trademark that sets Deaf people apart from the communication behavior of all other groups of people. It is the reason we say that Deaf people represent a linguistic minority. It is also why some people who are deaf do not see themselves as belonging to the Deaf community. 
% We use the lowercase spelling of deaf to refer to a person or a group of people who have a substantial degree of hearing loss. Having a hearing loss does not mean that a person automatically knows how to sign. If a deaf person does not know sign language, then that person will not be able to access the varied cultural experiences associated with the Deaf community. Communication is basic, and ASL is the communication of the Deaf community.

% Can a nondeaf person who is fluent in ASL be a member of the Deaf community? No. Deaf people do not view nondeaf people as members of their community because nondeaf people lack a third critical characteristic, which is shared experiences. If you have normal hearing, then you will never have the experiences of a life that is centered on seeing. 

% Let's put all of this in perspective. Let's say you have a young friend who acquired a hearing loss and was fitted with hearing aids. WOuld we say that he was Deaf? No, we would say that he is hard of hearing because speaking is still his main means of communication. If his hearing continues to deteriorate, then we might say that he is becoming deaf; that is, he is acquiring a substantial degree of hearing loss. What if his difficulty with hearing leads him to learn to sign ASL, which becomes his primary language of communication, and he begins participating in some activities in the Deaf community? Would we say then that he is Deaf? We would probably say, "friend, welcome to the club".


% Technology and Other Adaptations
% "Many of the technological advances for the majority in our society have penalized Deaf people. This irony emerges most clearly in telecommunications. The invention of the telephone made it difficult for Deaf people to compete in the labor market. Radio became an important means of broadcasting information, whether commercial, political, governmental, or whatever, further cutting off Deaf people from the larger society surrounding them. Television did little to improve the situation, though it embraced the technology that could have (and to some extent does) include Deaf people. Talking pictures were a blow to the entertainment and education of Deaf people; they could enjoy the "silents" on par with the rest of the audience. But Deaf people and their supporters have not passively accepted the status quo. They have taken steps to reduce the handicap the new technologies have imposed". -- Jerome Schein (At Home Among Strangers)

% -----
% ASL Grammar
% ASL has visual, spatial, and gestural features that combine to create some grammatical structures that are unparalleled in the world of spoken languages. As you learn about ASL grammar, look for similarities in the world of spoken languages. [...]
% In particular, the spatial qualities of ASL grammar allow a signer to express more than one thought simultaneously -- a characteristic that cannot be duplicated in English. Facial grammar, or nonmanual signals, is also a significant component of ASL.

% Facial Grammar
% What's in a sign may not be what's in the mind. To capture the sense of what a signer is signing, you must read the signer's face and body. When you listen to someone speak, you listen not only to the words but also to how the words are spoken. The tone of the voice, the rise and fall of the pitch, the length of the pause, and the steadiness of voice are all features that you latch onto with little effort in your spoken communication.
% These traits are nonexistent in signing, but they do have parallel traits that are crucial to ASL's grammar. The raised eyebrow, the tilted head, the open mouth, hunching of the shoulders, and a sign held slightly longer than others shape the meaning of the signs that are made by the hands. We call these nonmanual signals (NMS) facial grammar, which allows you to use facial expressions, your body, and gestures to add meaning and additional information to your signing. Mastering ASL cannot occur without a mastery of facial grammar.


% ===================================================
% \cite{hill-2019-sign-languages}
% ===================================================
%
% [Pág 2]
% Sign languages are produced by the hands, face, and body and perceived primarily visually, in contrast to spoken languages, which are produced by the mouth and vocal tract and perceived primarily auditorily (although manual gestures and visual perception of gestures and mouth movements are also important for spoken languages). Natural sign languages emerge (are not invented) when Deaf people form a community, often through educational systems. Sign languages are, therefore, primarily the languages of Deaf people, who cherish them for their cultural and community-building value. 
% It is important to recognize the connection between sign languages and Deaf communities. Until relatively recently, Deaf communities have been told (explicitly and implicitly) that their “sign communication” was inferior, broken, unimportant, or insufficient. Educational systems and the broader hearing majority community would stress the value of learning the spoken language, even at the expense of the sign language. In fact, such attitudes persist, both in areas where the national sign language has not been deeply studied linguistically and in areas where it has been studied but the focus for economic advancement is on the spoken language. However, the natural sign languages of Deaf communities are completely linguistic, rule-governed, capable of expressing anything, and fully worthwhile. We unreservedly endorse such affirmations of the value of sign languages and promote their use in all aspects of the lives of Deaf people. 

% Who belongs to the Deaf community? The “d” is capitalized to reinforce the view that Deaf communities form cultural groups with practices and values that are in some cases distinct from those of non-Deaf communities. These cultural effects are passed down within the community, from parents to children in some cases, but more often through interactions of Deaf people from different families. The leaders of Deaf communities are usually Deaf adults who were raised with Deaf parents or within the community from a very early age. Generally, members of the Deaf community are audiologically deaf or hard-of-hearing (and they shun the label “hearing impaired”). The hearing children born to Deaf parents are often known as Codas (from the name of an organization, CODA, ‘children of Deaf adults’), and they are sometimes part of the Deaf community. 

% It is important to note that people have many identities with intersectional effects, and in this respect, not all Deaf people have the same experiences, values, and life view. A Deaf person’s identity as Deaf will be affected by their identity in other ways, including race, ethnicity, gender identity, etc. Almost all research on the American Deaf community has focused only on a subset of Deaf people, so it is important to bear in mind that others might share some but not all of the characteristics described here. 
% Sign languages are, then, Deaf languages. Just as with the languages of other minority groups who have experienced oppression, hearing researchers who benefit from the study of sign languages (both in personal satisfaction and in economic, career, and other means) must acknowledge the primacy of Deaf signers and treat their language with the utmost respect.

%---
% Phonology
% Phonology is often described as the study of the sounds of language and their organization. If that is the way to look at phonology, it would be appropriate to say that sign languages do not have phonology. However, phonology is actually more abstract. It is about the ways in which words are made up of pieces that are not meaningful. It is about what these pieces are and how they work together. In spoken languages, the actual sounds that make up words are part of the study of phonology; yet, phonology is more concerned with the ways we can define these component pieces, how they change in different contexts (e.g., different words), and how they are organized.
% If there are component pieces that make up individual signs, then there is sign phonology. If there are implicit rules about the ways that the pieces can and cannot combine, then there is phonology. Indeed, one of the first linguistic discoveries about American Sign Language (ASL) is that “signs have parts” – [...]

% Morphology
% Morphology is the branch of linguistics that studies how words are formed from component parts. A morpheme is generally described as a consistent pairing of form (e.g., a sequence of sounds or a combination of handshape, location, and movement) and meaning, but there are morphemes that change their form in different contexts as well as those that don’t seem to have a consistent meaning.



% ===================================================
% \cite{quadros-2004-estudos-linguisticos}
% ===================================================
% 
% A linguística é o estudo científico das línguas naturais e humanas.
% ---
% A linguística busca desvendar os princípios independentes da lógica e da informação que determinam a linguagem humana. Tais princípios são o que há de comum nos seres  humanos que possibilitam a realização das diferentes línguas. Portanto, nesse  sentido, a teoria linguística extrapola as questões do uso.  A área da linguística está crescendo como área de estudo,
% ---
% A linguística é uma ciência que busca respostas para problemas essenciais relacionados à linguagem  que precisam ser explicados: Qual a natureza da linguagem humana? Como  a comunicação se constitui? Quais os princípios que determinam a habilidade dos seres humanos em produzir e compreender a linguagem? A linguística desmembra tais questões com a finalidade de explicar os problemas e  elaborar uma teoria da linguagem humana e uma teoria da comunicação. 
%
% ---
% Stokoe, em 1960,  percebeu e comprovou que a língua dos sinais atendia a todos os critérios  linguísticos de uma língua genuína, no léxico, na sintaxe e na capacidade de  gerar uma quantidade infinita de sentenças.  Stokoe observou que os sinais não eram imagens, mas símbolos abstratos complexos, com uma complexa estrutura interior. Ele foi o primeiro, portanto, a procurar uma estrutura, a analisar os sinais, dissecá-los e a pesquisar  suas partes constituintes. Comprovou, inicialmente, que cada sinal apresentava pelo menos três partes independentes (em analogia com os fonemas da  fala) – a localização, a configuração de mãos e o movimento – e que cada parte possuía um número limitado de combinações.
% ---
% Naturalmente que o trabalho de Stokoe (1960) representou o primeiro passo em relação aos estudos das línguas de sinais. Pesquisas posteriores, feitas em grande parte com a língua de sinais americana, mostraram,  entre outras coisas, a riqueza de esquemas e combinações possíveis entre  os elementos formais que servem para ampliar consideravelmente o vocabulário básico. 

% Fonologia das línguas de sinais
% Fonologia das línguas de sinais é o ramo da lingüística que objetiva identificar a estrutura e a organização dos constituintes fonológicos, propondo  modelos descritivos e explanatórios. A primeira tarefa da fonologia para línguas de sinais é determinar quais são as unidades mínimas que formam os  sinais. A segunda tarefa é estabelecer quais são os padrões possíveis de combinação entre essas unidades e as variações possíveis no ambiente fonológico. 


% Morfologia das línguas de sinais
% Morfologia é o estudo da estrutura interna das palavras ou dos sinais,  assim como das regras que determinam a formação das palavras. A palavra  morfema deriva do grego morphé, que significa forma. Os morfemas são as  unidades mínimas de significado.  
% Alguns morfemas por si só constituem palavras, outros nunca formam  palavras, apenas constituindo partes de palavras. Desta forma, têm-se os  morfemas presos que, em geral, são os sufixos e os prefixos, uma vez que não  podem ocorrer isoladamente, e os morfemas livres que constituem palavras.  
% Mas, na medida em que se pode formar palavras a partir de outras palavras, é importante reconhecer que as palavras podem ser unidades complexas, constituídas de mais de um elemento. 
% ---
% Assim como as palavras em todas as línguas humanas, mas diferentemente dos gestos, os sinais pertencem a categorias lexicais ou a classes de palavras  tais como nome, verbo, adjetivo, advérbio, etc. As línguas de sinais têm um  léxico e um sistema de criação de novos sinais em que as unidades mínimas  com significado (morfemas) são combinadas. Entretanto, as línguas de sinais  diferem das línguas orais no tipo de processos combinatórios que freqüentemente cria palavras morfologicamente complexas. Para as línguas orais, palavras complexas são muitas vezes formadas pela adição de um prefixo ou sufixo  a uma raiz. Nas línguas de sinais, essas formas resultam freqüentemente de  processos não-concatenativos em que uma raiz é enriquecida com vários movimentos e contornos no espaço de sinalização (Klima e Bellugi, 1979). 

% [...]





% ===================================================
% \cite{jay-2011-dont-just-sign}
% ===================================================
% To put it simply, grammar is a set of rules for using a language. The grammar of a language includes the phonology, morphology, and syntax of that language. 
% A language’s grammar is determined by the group of people who use that language. The main users of American Sign Language are the members of the Deaf community. The way that the Deaf community uses ASL is what constitutes ASL grammar. 
% American Sign Language has its own grammar system. This means that ASL has its own rules for phonology, morphology, and syntax.

% Phonology
% In language, phonology is the study of the smallest part of the language that conveys meaning. In spoken languages, like English, a phoneme is a unit of sound that conveys meaning. 
% In ASL, the smallest parts of the language, the phonemes, are handshape, movement, palm orientation, location, and facial expression. If you change any of these parameters of a sign, then you have changed the meaning of the sign.

% The Five Sign Parameters
% Just like how we see English words as the arrangement of letters, there are five basic sign language elements that make up each sign. If any of these parameters are changed when creating a sign, the meaning of the sign can change. 
% The first four elements are: 
% • Handshape – This is the shape of your hand that is used to create the sign. 
% • Movement – This is the movement of the handshape that makes the sign. 
% • Palm orientation – This is the orientation of your palm when making the sign. 
% • Location – This is the location of the sign on or in front of your body. 
% There is also a fifth element that has recently been included with this list: 
% • Non-manual Markers – This is the various facial expressions or body movements that are used to create meaning. 
% American Sign Language is a very expressive language, and understanding these elements will give you a better understanding of how signs are made and what makes them different.


% Morphology
% In language, morphology is the study of the forms and formations of words. A morpheme is the smallest indivisible unit of syntax that retains meaning.
%For example, in English, the word “threateningly” consists of four morphemes: “threat,” which is a noun; “en,” which changes the noun into a verb; “ing,” which changes it into an adjective; and “ly” which changes it into an adverb. 
% In ASL, there are no signs for affixes like “en,” “ing,” “ly,” etc. to change the meaning of words. Instead, ASL uses non-manual markers, changes in parameters, and other signs to indicate tense, degree, intensity, plurality, aspect, and more.

% • Inflection (Adverbs)
% • Noun-Verb Pairs
% • Classifiers
% • Verbs
% • Time



